A Simple Type Theory for Category Theory

### 2022-04-21 paper/paper.tex
category of elements, category with families? and start of GAT semantics

@@

\subsection{Graph of a Profunctor}

One final multiplicative we can define is the \emph{graph} of a
profunctor. Given a profunctor $\alpha:\cat C;\beta:\cat D \vdash R$,
the ``graph'' of $R$ is a category $\exists \alpha;\beta. R$. In
ordinary category theory, this is usually defined as the category of
elements of $R$ viewed as a functor $R : \cat C^o \times D \to \Set$.
In detail, its objects are triples $(a:\cat C, r : R, b : \cat D)$ and
morphism from $(a,r,b)$ to $(a', r', b')$ consists of a pair of
morphisms $g : b \to b'$ and a backwards $f : a' \to a$ such that $r'
= g\circ r \circ f$.
%
By picking $R = \punitinXfromYtoZ {\cat C} {\alpha}{\alpha'}$ we
recover the \emph{arrow} category $\exists
(\alpha;\alpha'). \punitinXfromYtoZ {\cat C}{\alpha}{\alpha'}$, by
composing these with functors, we recover the notion of \emph{comma}
category $\exists \alpha;\beta. \punitinXfromYtoZ {\cat
  E}{(F\alpha)}{(G\beta)}$ and by picking either $F$ or $G$ to have
domain the terminal category $1$, we recover the notion of a
\emph{slice} category.

@@

@@

@@

@@

Next, we develop the basic model theory for SCT. Because SCT includes
not just a syntax for category theory, but is also indexed by an
ordinary context of dependent type theory $\Gamma$, a model of SCT is
not given by a \emph{single} virtual double category, but instead by
an \emph{indexed family} of VDCs, parameterized by an object of a
category that interprets dependent type theory. We call such a
structure as a \emph{VDC-hyperdoctrine}. We then show how to construct
a VDC-hyperdoctrine indexed by the category of sets, which has the
intuitive reading that any of the VDC judgments $\Gamma \vdash J$ is
interpreted as simply a family of $J$-data parameterized by elements
of the set $\Gamma$.

@@

\begin{definition}{Indexed Virtual Equipment}
  A VE-hyperdoctrine is a pair of a category $C$ and a functor $\cat
  C^{op} \to VE$. A morphism of VE-hyperdoctrines
  \[(F,\phi) : (\cat C, \mathcal V : \cat C^{o} \to VE) \to (\cat D, \mathcal U : \cat D^{o} \to VE)\]
  is a pair of a functor $F : \cat C \to \cat D$ and a natural
  transformation $\phi : \mathcal V \Rightarrow\mathcal U \circ F^{o}$.
\end{definition}

\begin{definition}{Soundness}

\end{definition}

\begin{definition}{Completeness}
  We define the syntactic VE-hyperdoctrine $\textrm{Syn}$ as follows:
  \begin{enumerate}
  \item The base category is the category of contexts $\Gamma$ and substitutions.
  \item Define $\mathcal V_\Gamma$ as follows
    \begin{enumerate}
    \item An object of $\mathcal V^\Gamma$ is a proof $\Gamma \vdash \cat C \isaCat$.
    \item A vertical morphism $\mathcal V^\Gamma(\cat C, \cat D)$ is a
      term $\Gamma\pipe \alpha:\cat C \vdash t : \cat D$ moduleo
      $\alpha$-equivalence. Composition and identity are given by composition/variable.
    \item A horizontal morphism $R \in \mathcal^\Gamma_h$ is a term
      $\Gamma\pipe \alpha:\cat C;\beta:\cat D \vdash R$ with $d^-R =
      \cat C$ and $d^+R = \cat D$.
    \item Restriction of $R$ along $s,t$ is given by substitution
      $R[s/\alpha;t/\beta]$. This is clearly associative and unital.
    \item A 2-cell from $\Phi$ to $A$ is a term $\Phi \vdash t : B$
    \end{enumerate}
  \end{enumerate}
\end{definition}

@@

And we define a profunctor of interpretations of a signature in a
virtual equipment as follows:
\begin{definition}
  An interpretation of a signature $\Gamma$ in a virtual equipment
  $\mathcal V$ is an element of the set $\sem{\Gamma}_{\mathcal
    V}$. Call this set $\textrm{Interp}(\Gamma;\mathcal V)$.

  Note that this constitutes a profunctor contravariant in $\Gamma$
  and covariant in $\mathcal V$.
\end{definition}

Next, we define the syntactic virtual equipment $\textrm{Syn}(\Gamma)$
for a signature $\Gamma$.
\begin{definition}
  For any $\Gamma \isadtctx$, we construct a virtual double category
  $V^\Gamma$.
  \begin{enumerate}
  \item $V^\Gamma_v$ is the set of terms $\Gamma \vdash \cat C \isaCat$.
  \item $V^\Gamma_v(\cat C, \cat D)$ is the set of objects in context
    $\Gamma \pipe \alpha:\cat C \vdash a : \cat D$. The category
    structure is given by substitution, with the variable $\alpha
    \vdash \alpha$ the identity.
  \item $V_h$ is the set of triples $(\cat C, \cat D, \Gamma \pipe
    \alpha:\cat C;\beta:\cat D \vdash R \isaSet)$ where $d^\pm$ are
    the projections
  \item $V^\Gamma_h(\vec R; S)$ is the set of triples $(a,b,\Gamma
    \pipe \Phi \vdash t : S[a;b])$ where $\Phi$ is the context
    corresponding to $\vec R$ and $d^\pm$ are the obvious projections.

  \item The identity morphisms are $(\cat C, \cat D, R)$ is $\Gamma\pipe\alpha:\cat C, x:R\beta:\cat D \vdash x:R$
  \item Composition $(a,b,t) \circ [(a_1,),\ldots]$ is given by
    substitution in each component: $(a[a_1],b[a_n], t[\phi])$.
  \item Associativity and unit rules for composition follow from the
    definition of substitution by a straightforward induction.
  \end{enumerate}
\end{definition}

Then the soundness theorem says that for any $\Gamma$, the virtual
double category $\textrm{Syn}(\Gamma)$ provides an isomorphism of profunctors:
\[ \textrm{Interp}(\Gamma, \mathcal V \cong VE(\textrm{Syn}(\Gamma), \mathcal V) \]
As we saw in a previous section, the backwards direction of this
isomorphism should be a universal interpretation:

\begin{definition}[Universal Interpretation]
  For any signature $\Gamma$ there is a ``self-interpretation'' $\eta:
  \textrm{Interp}(\Gamma, \textrm{Syn}(\Gamma))$ defined as the
  identity substitution $\id \in \sem{\Gamma}_{\textrm{Syn}(\Gamma)}$

\end{definition}

\section{Example Models}

@@

  \begin{enumerate}
  \item Objects are functions $X \to \mathcal V_0$.
  \item Vertical morphisms $\mathcal V^X_v(A,B)$ are functions
    $\Pi_{x\inX} V_v(A x,B x)$. With pointwise composition, identity.
  \item Horizontal morphisms $\mathcal V^X_h$ are functions $\Gamma \to \mathcal V_h$, with
    $(d^\pm R)x = d^\pm (Rx)$.
  \item Restriction of $R$ along $f,g$ is given by $(R(f,g))x = (R x)(f x,g x)$.
  \item A 2-cell from $\Phi$ to $R$ is a family $\Pi_{x\in X} \mathcal
    V^X_h(\Phix,R x)$ with composition given pointwise.
  \end{enumerate}
\end{definition}

@@

  \begin{enumerate}
  \item The indexing cateogry is $\Set$.
  \item The equipment is defined as $\mathcal U\Gamma = \mathcam V^\Gamma$.
  \item Restriction by a function $\gamma : \Delta \to \Gamma$ is
    given pointwise.
  \end{enumerate}
\end{definition}

>>>>

\newcommand{\paramPresheaf}[1]{\mathcal P^{#1}}
\newcommand{\pmPresheaf}{\paramPresheaf\pm}
\newcommand{\posPresheaf}{\paramPresheaf+}
\newcommand{\negPresheaf}{\paramPresheaf-}
\newcommand{\negPresheafApp}[2]{#1\,#2}
\newcommand{\posPresheafApp}[2]{#1\,#2}

@@

\subsection{Category of Elements}

One final multiplicative we can define is the \emph{category of
elements} of a profunctor. Given a profunctor $\alpha:\cat
C;\beta:\cat D \vdash R$, the ``category of elements'' of $R$ is a
category $\Sigma_{\alpha:\cat C;\beta:\cat D}. R$. In detail, its
objects are triples $(a:\cat C, r : R, b : \cat D)$ and morphism from
$(a,r,b)$ to $(a', r', b')$ consists of a pair of morphisms $g : b \to
b'$ and a backwards $f : a' \to a$ such that $r' = g\circ r \circ f$.
%
By picking $R = \punitinXfromYtoZ {\cat C} {\alpha}{\alpha'}$ we
recover the \emph{arrow} category $\exists
(\alpha;\alpha'). \punitinXfromYtoZ {\cat C}{\alpha}{\alpha'}$, by
composing these with functors, we recover the notion of \emph{comma}
category $\exists \alpha;\beta. \punitinXfromYtoZ {\cat
  E}{(F\alpha)}{(G\beta)}$ and by picking either $F$ or $G$ to have
domain the terminal category $1$, we recover the notion of a
\emph{slice} category.

@@

\subsection{Positive and Negative Presheaf Categories}

Though defining arbitrary functor categories type-theoretically would
require a ``multi-argument functor'' judgment, we can define
\emph{presheaf} categories by the universal property that a functor
$\cat C \to \mathcal P {\cat D}$ is the same as a profunctor from
$\cat D$ to $\cat C$. To distinguish the covariant from contravariant
presheaves, we define them dually as $\negPresheaf {\cat C}$ and
$\posPresheaf {\cat D}$

\begin{figure}
  \begin{mathpar}
    \inferrule*[right=NegPresheaf-Formation]
    {\Gamma \vdash \cat C \isaCat}
    {\Gamma \vdash \negPresheaf {\cat C} \isaCat}

    \inferrule*[right=NegPresheaf-E]
    {\Gamma \pipe d^-\Xi \vdash a : \cat C \and
     \Gamma \pipe d^+\Xi \vdash p : \negPresheaf {\cat C}
    }
    {\Gamma \pipe \Xi \vdash \negPresheafApp a p \isaSet}

    \inferrule*[right=NegPresheaf-I]
    {\Gamma \pipe \alpha:\cat C;\beta:\cat D \vdash R : \isaSet}
    {\Gamma \pipe \beta:\cat D \vdash \lambda \alpha:\cat C. R : \negPresheaf {\cat C}}

    \negPresheafApp a (\lambda \alpha. R) = R[a/\alpha]\\

    \inferrule*[right=NegPresheaf-$\eta$]
    {\Gamma \pipe \beta:\cat D \vdash p : \negPresheaf {\cat C}}
    {p = \lambda \alpha. \negPresheafApp \alpha p}

    \inferrule*[right=PosPresheaf-Formation]
    {\Gamma \vdash \cat D \isaCat}
    {\Gamma \vdash \posPresheaf {\cat D} \isaCat}

    \inferrule*[right=PosPresheaf-E]
    {\Gamma \pipe d^-\Xi \vdash p : \posPresheaf {\cat D}\and
     \Gamma \pipe d^+\Xi \vdash a : \cat D
    }
    {\Gamma \pipe \Xi \vdash \posPresheafApp p a \isaSet}

    \inferrule*[right=PosPresheaf-I]
    {\Gamma \pipe \alpha:\cat C;\beta:\cat D \vdash R : \isaSet}
    {\Gamma \pipe \alpha:\cat C \vdash \lambda \beta:\cat D. R : \posPresheaf {\cat D}}

    \posPresheafApp {(\lambda \beta. R)} b = R[b/\beta]\\

    \inferrule*[right=PosPresheaf-$\eta$]
    {\Gamma \pipe \alpha:\cat C \vdash p : \posPresheaf {\cat D}}
    {p = \lambda \beta. \posPresheafApp p \beta}
  \end{mathpar}
  \caption{Presheaf Categories}
\end{figure}

@@

\subsection{Yoneda Lemma and Representable Presheaves}

Given any category $\cat C$, we can define the positive and negative
Yoneda embeddings $Y^\pm : \cat C \to \pmPresheaf {\cat C}$ using the
hom profunctor. Note that both of these functors are covariant in
$\cat C$, so semantically, $\posPresheaf {\cat C}$ should be
interpreted as $(\Set^{\cat C})^o$.

\begin{definition}
  Define the negative Yoneda embedding $Y^-[\alpha]$
  \[ \alpha:\cat C \vdash Y^-[\alpha] = \lambda \alpha':\cat C. \punitinXfromYtoZ {\cat C}{\alpha'}{\alpha} : \negPresheaf {\cat C} \]

  The positive Yoneda embedding is defined similarly $Y^+[\alpha]$.
\end{definition}

The Yoneda lemma says that the set of natural transformations from
$Y^-[\alpha]$ to a presheaf $p$ is naturally isomorphic to the
elements of the presheaf itself: $\negPresheafApp \alpha p$. In
encoding this statement in our type theory, we find that we can use
two different interpretations of ``natural transformations''.

On the one hand, a natural transformation should be a morphism in the
presheaf category. This interpretation of the Yoneda lemma would mean
that we have a natural isomorphism of profunctors:
\[ \alpha:\cat C; \pi:\negPresheaf {\cat C} \vdash \punitinXfromYtoZ{\negPresheaf {\cat C}} {Y^-[\alpha]} {\pi} \cong \negPresheafApp \alpha \pi \]
Attempting to prove this we become stuck very quickly because we know
nothing about the structure of morphisms in the presheaf category.

On the other hand, a natural transformation of presheaves $\pi$ and
$\rho$ can be defined as a family of functions from $\negPresheafApp
\alpha \pi$ to $\negPresheafApp \alpha \rho$, or in other words as an
\emph{end}. Specializing to the Yoneda lemma, we interpret ``a natural
transformation from $Y^-[\alpha]$ to $\pi$'' as the following, which simplifies using $\beta$ reduction:
\[ \alpha:\cat C; \pi:\negPresheaf {\cat C} \vdash \homrallXYtoZ {\alpha':\cat C} {\negPresheafApp {\alpha'} {Y^-[\alpha]}}{\negPresheafApp {\alpha'} \pi} = \homrallXYtoZ {\alpha':\cat C} {(\punitinXfromYtoZ {\cat C}{\alpha'}{\alpha})}{(\negPresheafApp {\alpha'} \pi)} \]
And then the Yoneda lemma becomes:
\[ \alpha:\cat C; \pi:\negPresheaf {\cat C} \vdash \homrallXYtoZ {\alpha':\cat C} {(\punitinXfromYtoZ {\cat C}{\alpha'}{\alpha})}{(\negPresheafApp {\alpha'} \pi)} \cong \negPresheafApp \alpha \pi\]
Which is an instance of our earlier Lemma\,\ref{lem:yo-and-co-yo}
which follows quite easily.

Then we are left with the question of whether these two forms of the
Yoneda lemma are equivalent, or slightly more generally are these two definitions of natural transformation of presheaves equivalent:
\[ \rho:\negPresheaf{\cat C}; \pi:\negPresheaf {\cat C} \vdash \punitinXfromYtoZ{\negPresheaf {\cat C}} {\rho} {\pi} \]
and
\[ \rho:\negPresheaf{\cat C}; \pi:\negPresheaf {\cat C} \vdash \homrallXYtoZ {\alpha:\cat C} {\negPresheafApp \alpha \rho}{\negPresheafApp \alpha \pi} \]
While we can show that there is a morphism from the hom set to the
end, the inverse does not seem provable, and seems analogous to the
\emph{univalence} axiom in homotopy type theory, as it tells us that
the morphisms of the universe are given by a particular kind of
function in our logic.

@@

\subsection{Formal Type Theory}

\begin{definition}[Category with Families]
  A category with families consists of
  \begin{enumerate}
  \item A category $\textrm{Ctx}\isaCat$
  \item A presheaf $\textrm{Ty} : \harr {\textrm{Ctx}}{1}$
  \item A presheaf $\textrm{Tm} : \harr {\Sigma_{\Gamma:\textrm{Ctx};*:1}\harrapp{\textrm{Ty}} \Gamma *} 1$
  \item A terminal object in $\textrm{Ctx}$
  \item An \emph{extension operation}
    $\textrm{ext} : \Sigma_{\Gamma:\textrm{Ctx};*:1}\harrapp{\textrm{Ty}}\Gamma * \to \textrm{Ctx}$
    right representing the profunctor
    \[ \Delta:\textrm{Ctx};A:\Sigma_{\Gamma:\textrm{Ctx};*:1}\harrapp{\textrm{Ty}}{\Gamma}{*} \vdash \tensorexistsXwithYandZ {a}{(\punitinXfromYtoZ {} \Delta {\pi^-\pi^-a})}
       {(\punitinXfromYtoZ {}{\pi^-a}{A})}\]
       % {\Sigma_{\Gamma:\textrm{Ctx};*:1}\harrapp{\textrm{Ty}}{\Gamma}{*}}
    % 
  \end{enumerate}
\end{definition}

@@

Next, we develop the basic model theory for SCT. Because SCT includes
not just a syntax for category theory, but is also indexed by an
ordinary context of dependent type theory $\Gamma$, a model of SCT is
not given by a \emph{single} virtual double category, but instead by
an \emph{indexed family} of VDCs, parameterized by an object of a
category that interprets dependent type theory. We call such a
structure a \emph{VDC-hyperdoctrine}. We then show how to construct a
from any VDC, a VDC-hyperdoctrine indexed by the category of sets,
which has the intuitive reading that any of the VDC judgments $\Gamma
\vdash J$ is interpreted as simply a family of $J$-data parameterized
by elements of the set $\Gamma$.

@@

\begin{definition}{Indexed Virtual Equipment}
  A VE-hyperdoctrine is a pair of a category with families $(C,
  \textrm{Ty}, \textrm{Tm})$ and a functor $\cat C^{op} \to VE$. A
  morphism of VE-hyperdoctrines
  \[(F,\phi) : (\cat C, \mathcal V : \cat C^{o} \to VE) \to (\cat D, \mathcal U : \cat D^{o} \to VE)\]
  is a pair of a functor of categories with families $F : \cat C \to
  \cat D$ and a natural transformation $\phi : \mathcal V
  \Rightarrow\mathcal U \circ F^{o}$.
\end{definition}

@@

\subsection{Generalized Algebraic Theory of VE Hyperdoctrines}

To simplify our initiality proof, we work with an \emph{essentially
algebraic} presentation of the notion of VE hyperdoctrine. As with all
GATs, this has an initial model, which is a kind of ``combinator''
syntax with explicit substitution for VE hyperdoctrines. Then we show
that our type theoretic syntax is equivalent to the combinator syntax
by a simple inductive ``simplification'' algorithm. We conjecture that
a full normalization proof is possible, which would establish useful
conservativity results for the calculus.

The theory has sorts Ctx, Subst, Tm, Ty, Cat, Fun, Prof, ProfCtx and
Trans.

\subsection{Example Models}

@@

  \begin{enumerate}
  \item Objects are functions $X \to \mathcal V_0$.
  \item Vertical morphisms $\mathcal V^X_v(A,B)$ are functions
    $\Pi_{x\in X} V_v(A x,B x)$. With pointwise composition, identity.
  \item Horizontal morphisms $\mathcal V^X_h$ are functions $\Gamma \to \mathcal V_h$, with
    $(d^\pm R)x = d^\pm (Rx)$.
  \item Restriction of $R$ along $f,g$ is given by $(R(f,g))x = (R x)(f x,g x)$.
  \item A 2-cell from $\Phi$ to $R$ is a family $\Pi_{x\in X} \mathcal
    V^X_h(\Phi x,R x)$ with composition given pointwise.
  \end{enumerate}
\end{definition}

@@

  \begin{enumerate}
  \item The indexing cateogry is $\Set$.
  \item The equipment is defined as $\mathcal U\Gamma = \mathcal V^\Gamma$.
  \item Restriction by a function $\gamma : \Delta \to \Gamma$ is
    given pointwise.
  \end{enumerate}
\end{definition}
